A Dependently Typed Multi-Stage Calculus

### 2019-08-29 text/formal-defnition.tex
Describe the motivation more precisely.
Revise the vadd example.

We show an example of a dependently typed code generator in a
hypothetical language based on \LTP.  
% This rule is motivated by a practical consideration.
Let's consider an application, for example, in computer graphics, in
which we have potentially many pairs of vectors of the fixed (but
statically unkown) length and a function---such as vector
addition---to be applied to all of them.

\AI{Some more description about what we'd like to do.}
\AI{Brief comparison with other ways (somewhere else)?}
% \AK{I strenghened this example by adding concrete situation. Do we need
% more reinforcements such as comparision with other method?} We will discuss
% the motivation through an example of specializing a vector addition function
% with respect to the size of vectors.

The function vadd (with an auxiliary
function vadd$_1$) below implements a code generator.
This language provides definitions by \textbf{let},
recursive functions (represented by \textbf{fix}), \textbf{if}-expressions,
and primitives cons, head, and tail to manipulate vectors. We assume that
$\text{cons}$ is of type $\Pi n:\I.\I \to \text{Vector}\ n \to \text{Vector}\ (n+1)$, 
$\text{head}$ is of type $\Pi n:\I.\text{Vector}\ (n+1) \to \I$, and
$\text{tail}$ is of type $\Pi n:\I.\text{Vector}\ (n+1) \to (\text{Vector}\ n)$.

The helper function vadd$_1$ takes a stage, the length $n$ of vectors, and two quoted vectors and returns code that computes the addition of the given two vectors:
%
\newcommand{\Vpn}{\text{Vector}\ (\%_\alpha n)}
\begin{tabbing}
	  $\textbf{let}\ \text{vadd}_1 : \F\alpha.\Pi n:\I.\TW_\alpha\Vpn\to\TW_\alpha\Vpn\to\TW_\alpha\Vpn$                                \\
	  \hspace{6mm} \= $= \textbf{fix}\ f.\Lambda\alpha.\lambda n:\I.\ \lambda v_1:\TW_\alpha\Vpn.\ \lambda v_2:\TW_\alpha\Vpn.$            \\
	  \> \hspace{6mm} \= $\textbf{if}\ n = 0 \ \textbf{then}\ \TB_\alpha \text{nil}$ \\
	  \>\> $\textbf{else}\ \TB_\alpha ($ \= $\textbf{let}\ t_1 = \text{tail}\ (\TBL_\alpha v_1)\ \textbf{in}$ \\
	  \>\>\> $\textbf{let}\ t_2 = \text{tail}\ (\TBL_\alpha v_2)\ \textbf{in}$ \\
          \>\>\> $\text{cons}\ (\text{head}\ (\TBL_\alpha v_1) + \text{head}\ (\TBL_\alpha v_2))\ (\TBL_\alpha f\ (n-1)\ t_1\ t_2))$
\end{tabbing}
(Here, we assume that the type system can determine whether $n=0$ when
\textbf{then}- and \textbf{else}-branches are typechecked so that
both branches can be given type \(\TW_\alpha \text{Vector }n\).)

The main function vadd below takes the length $n$ and returns
($\beta$-closed) code of a function that takes two vectors and returns
their addition using vadd$_1$.
\renewcommand{\Vpn}{\text{Vector}\ (\%_\beta n)}
\begin{align*}
	  & \textbf{let}\ \text{vadd}: \Pi n:\I.\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)                \\ 
	  & \hspace{6mm} = \lambda n:\I.\Lambda\beta.\TB_\beta (\lambda v_1:\Vpn.\ \lambda v_2:\Vpn. \\
	  & \hspace{63mm} \TBL_\beta (\text{vadd}_1\ \beta\ n\ (\TB_\beta\ v_1)\ (\TB_\beta\ v_2))) 
\end{align*}
\renewcommand{\Vpn}{\text{Vector}\ (\%_\beta 5)}%
If this function is applied to a
(nonnegative) integer constant, say 5, it returns function code for adding
two vectors of size 5.  The type of vadd\ 5, obtained by
substituting 5 for $n$, is
$\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)$.
\renewcommand{\Vpn}{\text{Vector}\ 5}
If the obtained code is run by applying to \(\varepsilon\),
the type of vadd\ 5\ $\varepsilon$ is
\(\Vpn\to\Vpn\to\Vpn\) as expected.

\AI{Comparison?}

\paragraph{Remark:}
When this function is composed with another piece of code $v$ of type, say,
\(\TW_\gamma \text{Vector }5\), \QPercent plays an essential role; that is,
\(\text{Vector }5\) and \(\text{Vector }(\%_\beta 5)\) are syntactically
different types but \QPercent enables to equate one with another.
Interestingly, Hanada and Igarashi~\cite{Hanada2014} rejected the idea of
reduction that removes $\%_\alpha$ when they developed \LTP{}, as such
reduction does not match the operational behavior of the CSP operator in
implementations. However, as an equational system for multi-stage programs,
the rule \QPercent makes sense.

\renewcommand{\Vpn}{\text{Vector}\ n}
\paragraph{Remark:} By using implicit type-level CSP, the type of vadd
could have been written
\(\Pi n:\I.\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)\).  In this type,
Vector\ $n$ is given kind at stage \(\varepsilon\) and type-level CSP
implicitly lifts it to stage \(\beta\).  However, if a type-level
constant takes two or more arguments from different stages, term-level
CSP is necessary.  A matrix type (indexed by the numbers of columns
and rows) would be such an example.

>>>>

We show an example of a dependently typed code generator in a
hypothetical language based on \LTP.  
% This rule is motivated by a practical consideration.
This language provides definitions by \textbf{let},
recursive functions (represented by \textbf{fix}), \textbf{if}-expressions,
and primitives cons, head, and tail to manipulate vectors. We assume that
$\text{cons}$ is of type $\Pi n:\I.\I \to \text{Vector}\ n \to \text{Vector}\ (n+1)$, 
$\text{head}$ is of type $\Pi n:\I.\text{Vector}\ (n+1) \to \I$, and
$\text{tail}$ is of type $\Pi n:\I.\text{Vector}\ (n+1) \to (\text{Vector}\ n)$.

Let's consider an application, for example, in computer graphics, in which we
have potentially many pairs of vectors of the fixed (but statically unkown)
length and a function---such as vector addition---to be applied to
them. This function should be fast because it is applied many times and be
safe because just one runtime error may ruin the whole long-running calculation.

% \AK{I strenghened this example by adding concrete situation. Do we need
% more reinforcements such as comparision with other method?} We will discuss
% the motivation through an example of specializing a vector addition function
% with respect to the size of vectors.

\newcommand{\Vpn}{\text{Vector}\ (\%_\alpha n)}

Our goal is to define the function vadd of type
\[
  \Pi n:\I.\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn).
\]
\renewcommand{\Vpn}{\text{Vector}\ n}
It takes the length $n$ and returns ($\beta$-closed) code of a
function to add two vectors of length $n$.  The generated
code is run by applying it to \(\varepsilon\) to obtain
a function of type \(\Vpn\to\Vpn\to\Vpn\) as expected.

We start with the helper function vadd$_1$, which takes a stage, the length $n$ of vectors, and two quoted vectors and returns code that computes the addition of the given two vectors:
%
\begin{tabbing}
	  $\textbf{let}\ \text{vadd}_1 : \F\alpha.\Pi n:\I.\TW_\alpha\Vpn\to\TW_\alpha\Vpn\to\TW_\alpha\Vpn$                                \\
	  \hspace{6mm} \= $= \textbf{fix}\ f.\Lambda\alpha.\lambda n:\I.\ \lambda v_1:\TW_\alpha\Vpn.\ \lambda v_2:\TW_\alpha\Vpn.$            \\
	  \> \hspace{6mm} \= $\textbf{if}\ n = 0 \ \textbf{then}\ \TB_\alpha \text{nil}$ \\
	  \>\> $\textbf{else}\ \TB_\alpha ($ \= $\textbf{let}\ t_1 = \text{tail}\ (\TBL_\alpha v_1)\ \textbf{in}$ \\
	  \>\>\> $\textbf{let}\ t_2 = \text{tail}\ (\TBL_\alpha v_2)\ \textbf{in}$ \\
          \>\>\> $\text{cons}\ $\= $(\text{head}\ (\TBL_\alpha v_1) + \text{head} \ (\TBL_\alpha v_2))$ \\
          \>\>\>\> $(\TBL_\alpha f\ (n-1)\ (\TB_\alpha t_1)\ (\TB_\alpha t_2)))$
\end{tabbing}
Note that the generated code will not contain branching on $n$ or recursion.
(Here, we assume that the type system can determine whether $n=0$ when
\textbf{then}- and \textbf{else}-branches are typechecked so that
both branches can be given type \(\TW_\alpha \text{Vector }n\).)

Using vadd$_1$, the main function vadd can be defined as follows:
\renewcommand{\Vpn}{\text{Vector}\ (\%_\beta n)}
\begin{align*}
	  & \textbf{let}\ \text{vadd}: \Pi n:\I.\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)                \\ 
	  & \hspace{6mm} = \lambda n:\I.\Lambda\beta.\TB_\beta (\lambda v_1:\Vpn.\ \lambda v_2:\Vpn. \\
	  & \hspace{63mm} \TBL_\beta (\text{vadd}_1\ \beta\ n\ (\TB_\beta\ v_1)\ (\TB_\beta\ v_2))) 
\end{align*}
\renewcommand{\Vpn}{\text{Vector}\ (\%_\beta 5)}%
The auxiliary function vadd$_1$ generates code to compute addition of
the formal arguments $v_1$ and $v_2$ without branching on $n$ or recursion.
As we mentioned already, if this function is applied to a
(nonnegative) integer constant, say 5, it returns function code for adding
two vectors of size 5.  The type of vadd\ 5, obtained by
substituting 5 for $n$, is
$\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)$.
\renewcommand{\Vpn}{\text{Vector}\ 5}
If the obtained code is run by applying to \(\varepsilon\),
the type of vadd\ 5\ $\varepsilon$ is
\(\Vpn\to\Vpn\to\Vpn\) as expected.

There are other ways to implement the vector addition function for
lists of fixed length: by using tuples instead of lists if the length
for all the vectors is statically known or by checking dynamically the
lengths of lists for every pair.  However, our method is better than
these two methods in two points.  First, our function, $\text{vadd}_1$
can generate functions for vectors of arbitrary length unlike the one
using tuples.  Second, $\text{vadd}_1$ has an advantage in speed over
the one using dynamic checking because it can generate an optimized
function for a given length.

We make two technical remarks before proceeding:
\begin{enumerate}
\item If the generated function code is composed with another piece of code of type, say,
\(\TW_\gamma \text{Vector }5\), \QPercent{} plays an essential role; that is,
\(\text{Vector }5\) and \(\text{Vector }(\%_\gamma 5)\), which would occur
by applying the generated code to \(\gamma\) (instead of \(\varepsilon\)), are syntactically
different types but \QPercent{} enables to equate them.
Interestingly, Hanada and Igarashi~\cite{Hanada2014} rejected the idea of
reduction that removes $\%_\alpha$ when they developed \LTP{}, as such
reduction does not match the operational behavior of the CSP operator in
implementations. However, as an equational system for multi-stage programs,
the rule \QPercent{} makes sense.
\item \renewcommand{\Vpn}{\text{Vector}\ n}
By using implicit type-level CSP, the type of vadd
could have been written
\(\Pi n:\I.\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)\).  In this type,
Vector\ $n$ is given kind at stage \(\varepsilon\) and type-level CSP
implicitly lifts it to stage \(\beta\).  However, if a type-level
constant takes two or more arguments from different stages, term-level
CSP is necessary.  A matrix type (indexed by the numbers of columns
and rows) would be such an example.
\end{enumerate}
